Compositional Quantitative Resource Analysis of C Programs

### 2014-06-26 paper/main.tex
(even) better into

Static analysis of quantitative properties of imperative programs is
an active research area and recent years have seen many innovations.
Notable tools that have been developed include
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.
%
While the these tools can derive impressive results for realistic
software, there still exist shortcomings that hamper the application
of the existing techniques in practice.
\begin{itemize}
\item Analysis tools are black boxes that either deliver a result or fail
  without enabling \emph{user interaction} to manually or
  semi-automatically derive bounds for challenging parts of the
  program.
\item The computed bounds are often \emph{non-compositional} local
  bounds (for a single (nested) loop) that are difficult to combine to
  global whole program bounds.
\item Existing techniques often rely on complex external tools such as
  abstract interpretation based invariant
  generation~\cite{GulwaniMC09} or translation of the program into a
  term-rewriting system~\cite{BrockschmidtEFFG14, SinnZV14} without
  providing \emph{verifiable certificates} for the correctness of the
  derived bound.
\end{itemize}
%
While there has been much progress in static quantitative analysis,
Knuth correctly points out in a recent interview~\cite{KnuthInter}
that the state-of-the-art in formal quantitative methods still falls
short in comparison with semantic techniques.
%
\begin{quote}
  % Consider, for example, a computer program that sorts a list of
  % numbers into order. 
  $[\cdots]$ Thanks to the work of Floyd, Hoare, and others,
  we have formal definitions of semantics, and tools by which we can
  verify that sorting is indeed always achieved. My job is to go
  beyond correctness, to an analysis of such things as the program's
  running time $[\cdots]$. I'm 100\% sure that my recurrence correctly
  describes the program's performance, and all of my colleagues agree
  with me that the recurrence is "obviously" valid. Yet I have no
  formal tools by which I can prove that my recurrence is right. I
  don't really understand my reasoning processes at all!
  \vspace{-3ex}
  \begin{flushright}
    -- Donald E. Knuth, 2014
  \end{flushright}
\end{quote}
%
In this work, we develop a resource analysis framework for C programs
that is based on a solid semantic foundation in which we start with a
formally defined cost model of C programs and derive resource bounds
together with easily verifiable certificates of their soundness with
respect to the cost model. The choice of C is primarily motivated by
our ongoing work on the formally verified hypervisor kernel
CertiKOS~\cite{GuVFSC11}.  CertiKOS is mainly developed in C and is
supposed to provide verified guaranties on timing and memory usage.
Moreover, C is a natural choice because it is still the most widely
used language for system development; in particular in safety-critical
embedded and real-time systems where resource bound analyses are often
required by regulatory authorities~\cite{May2013}.

Our starting point is a recent work~\cite{veristack14} in which we
have formally verified bounds on the stack usage of C programs.  A key
part of this stack-bound verification is based on a quantitative Hoare
logic for deriving abstract stack bounds for CompCert Clight programs
that depend on the depth of (recursive) function calls.  For
non-recursive programs it is possible to compute derivations for such
stack bounds in the quantitative Hoare logic automatically.  By
compiling these bounds together with the C program, we have been able
to automatically derive verified stack bounds for C programs without
recursion such as CertiKOS.

Our \emph{first contribution} is a generalization of the quantitative
Hoare logic for CompCert Clight that enables us to derive resource
bounds that are parametric in the resource of interest.  The
quantitative logic is implemented in the Coq Proof Assistant and
proved sound with respect to a small-step operational cost semantics
for Clight that counts a constant resource consumption for each atomic
execution step.  The semantics models the cost of terminating and
non-terminating executions.  The constant resource consumption at each
step is parametric in user-definable cost metric and can be negative
to express the release of resources (as needed, e.g., to reason about
memory usage).  Furthermore, a specific \emph{tick} function can be
inserted in the source code to further customize the cost model.

The idea of the quantitative Hoare logic is that we replace the
boolean predicates (in the pre- and postconditions) of Hoare logic
with numeric predicates that map a program state to a non-negative
number or infinity.  The intuitive meaning of a quantitative Hoare
triple $\htriple{P}{S}{Q}$ is that the program $S$ consumes less then
$P(\state)$ resources when executed in starting state $\state$, and
that at least $Q(\state')$ resources are left if $S$ terminates in
state $\state'$.  We call $P$ the \emph{potential} before the
execution of $S$ and $Q$ the potential after the execution of $S$.
Infinity corresponds to \emph{false} in classic Hoare logic and
boolean assertions can be embedded in the quantitative assertions to
express that a bound is only valid for a subset of inputs.  In the
implementation, we use a shallow embedding in Coq which makes the
logic very flexible.  There is practically no limitation on the format
of the derived resource bounds since they can be any function that is
definable in Coq.  The derived bounds can also be parametric in a set
of cost metrics or specifically apply to one fixed cost metric.

The precision and flexibility of the quantitative Hoare logic provide
a ample foundation for reasoning about resource consumption of Clight
programs.  The logic is naturally compositional and the quantitative
triples can be used to specify resource cost of library functions or
to describe resource contracts between different functions.  However,
reasoning about quantitative properties can often be more tedious then
reasoning about intentional properties.  Consequently, automation is
inevitable to derive resource bounds for large code bases like
CertiKOS.  Such an automation was easily achievable in the case of
constant stack bounds for a code base that does not contain recursive
functions.  In general however, resource bounds have to be parametric
in the arguments of a function and depend on the number of loop
iterations and (recursive) function calls that are performed by the
function.

For functional programs, there exists resource analysis systems that
are based on type-based amortized resource analysis~\cite{Jost03,
  HoffmannAH12} and can automatically derive complex polynomial bounds
with tight constant factors.  Type-based amortized resource analysis
has been our main inspiration in the design of the quantitative Hoare
logic and is therefore a natural candidate to automate the reasoning.
However, it is a well-known problem that type-based amortized resource
analysis works well for functional programs with pattern matching but
not for C-like programs with control flow that depends integer
arithmetic.  For example, it is a long-time open problem how to use
type-base amortized analysis to derive bounds for functions whose
resource consumption depends on (possibly negative) integers and
non-sequential control flow as introduced by \code{break} and
\code{return} statements.  Unfortunately, these are the functions with
non-constant resource consumption that are most common in CertiKOS and
other system software.

Our \emph{second contribution} is an automatic amortized resource
analysis for CompCert Clight that computes derivations in the
quantitative Hoare logic.  It is the first automatic amortized
resource analysis that can derive bounds that depend on negative
integers and handles programs with mutually recursive as well as
\code{break} and \code{return} statements.

XXXX

The main innovation that makes this work practical is our new
automatic amortized analysis for C code.  It targets programs that are
common in system software, that is, programs whose resource
consumption can be described as a function of sizes $|[x,y]|$ of
intervals of integer variables.  A typical bound would be $B(x,y,z) =
10{\cdot}|[x,y]| + 1{\cdot}|[z,0]|$ (for example $B(100,200,43) =
1043$).  The use of interval sizes in potential functions is an
important novelty.
%
Despite the apparent simplicity of the new analysis system, it is able
to reproduce results from the literature which were obtained using
methods based on abstract interpretation.  In contrast with abstract
interpretation--based methods, our technique does not require any
fixpoint computations on an abstract domain to obtain loop invariants.
The mechanism we designed is able to leverage local assertions such as
$x < y$ that we collect along the branching points of the program to
obtain global invariants.  We achieve this by generating a linear
constraint system that reflects resource cost and size changes of
integer variables in the program.  A solution of the linear program is
a resource usage bound for the C program.  We can naturally, handle
advanced control flow such as break, return, iteration over negative
integer variables, and mutually-recursive functions.

@@

We implemented our analysis for CompCert Clight, a subset of C in
which loops can only be exited with a \code{break} statement and
expressions are free of side effects.  C programs can for instance be
compiled to Clight programs using the verified CompCert C compiler.

XXXXXXXXX

In this article, we present the first resource analysis framework for
C that makes it possible to combine automatically and interactively
derived resource bounds.  It works automatically, efficiently and
precisely for many examples, can be used interactively for manual
bound derivation if the automation fails, and produces verifiable
certificates for the bounds.

XXXXXXX

In summary our contributions are as follows.
%
\begin{enumerate}
\item

\item We describe an automatic static analysis technique for deriving
  bounds for the cost that is defined by the operational semantics.
  It is based on amortized resource analysis for functional
  programs~\cite{Jost03,HoffmannAH12} and derives bounds that are
  non-negative linear combinations of sizes $|[x,y]|$ of intervals
  between program variables $x,y$.
\item We show with a publically available prototype implementation and
  an experimental evaluation that our novel amortized resource
  analysis works efficiently for realistic programs and that the
  derived constant factors in the bounds are close to or identical
  with the optimal ones.
\item We define a quantitative Hoare logic for interactively deriving
  cost bounds for C programs.  The logic is implemented and proved
  sound in the Coq Proof Assistant with respect to CompCert C
  light.  We show that our automatic analysis readily produces
  derivations in the quantitative Hoare logic.
% \item We prove the completeness of the program logic
%   with respect to the operational cost semantics.
\end{enumerate}

XXXXX

Our approach complements existing work since it provides a semantic
foundation for the computation of bounds while still providing support
for automation that sometimes goes beyond the capabilities of existing
techniques.  The quantitative Hoare logic and the automatic amortized
analysis enable semi-automatic interactive resource-bound development.
An automatically derived bound can be used in the quantitative logic
and a hand derived bound can be used by the automatic analysis.  Both
the automatic amortized analysis and the quantitative logic are
naturally compositional and describe the resource behavior of code
fragments without referring to the source code.  As a result, we
naturally derive global bounds that are functions of the input
parameters of the program.  To the best of our knowledge, our system
is the only one that can automatically handle resources like memory
that may become available during the execution.

XXXXX

@@

In this section, we informally introduce the quantiatative program
logic and the automatic amortized analysis for for Clight programs.
The purpose of this section is to build some intuition that should
make the following sections, which contain more details, easy to read.

>>>>

Static analysis of quantitative properties of imperative programs is
an active research area and recent years have seen many innovations.
Notable tools that have been developed include
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.
%
While the these tools can derive impressive results for realistic
software, there still exist shortcomings that hamper the application
of the existing techniques in practice.
\item Analysis tools are black boxes that either deliver a result or fail
  without enabling \emph{user interaction} to manually or
  semi-automatically derive bounds for challenging parts of the
  program.
\item The computed bounds are often \emph{non-compositional} local
  bounds (for a single (nested) loop) that are difficult to combine to
  global whole program bounds.
\item Existing techniques often rely on complex external tools such as
  abstract interpretation based invariant
  generation~\cite{GulwaniMC09} or translation of the program into a
  term-rewriting system~\cite{BrockschmidtEFFG14, SinnZV14} without
  providing \emph{verifiable certificates} for the correctness of the
  derived bound.
\end{itemize}
%
While there has been much progress in static quantitative analysis,
Knuth correctly points out in a recent interview~\cite{KnuthInter}
that the state-of-the-art in formal quantitative methods still falls
short in comparison with semantic techniques.
%
\begin{quote}
  % Consider, for example, a computer program that sorts a list of
  % numbers into order. 
  $[\cdots]$ Thanks to the work of Floyd, Hoare, and others,
  we have formal definitions of semantics, and tools by which we can
  verify that sorting is indeed always achieved. My job is to go
  beyond correctness, to an analysis of such things as the program's
  running time $[\cdots]$. I'm 100\% sure that my recurrence correctly
  describes the program's performance, and all of my colleagues agree
  with me that the recurrence is "obviously" valid. Yet I have no
  formal tools by which I can prove that my recurrence is right. I
  don't really understand my reasoning processes at all!
  \vspace{-3ex}
  \begin{flushright}
    -- Donald E. Knuth, 2014
  \end{flushright}
\end{quote}
%
In this work, we develop a resource analysis framework for C programs
that is based on a solid semantic foundation.  The choice of C is
primarily motivated by our ongoing work on the formally verified
hypervisor kernel CertiKOS~\cite{GuVFSC11}.  CertiKOS is mainly
developed in C and is supposed to provide verified guaranties on
timing and memory usage.  Moreover, C is a natural choice because it
is still the most widely used language for system development; in
particular in safety-critical embedded and real-time systems where
resource bound analyses are often required by regulatory
authorities~\cite{May2013}.  Our \emph{contributions} are as follows.
%
\begin{enumerate}
\item We define a operational cost semantics for CompCert Clight that
  defines the resource consumption of terminating and diverging
  executions.  The cost is parametric in a user-definable cost metric
  and can be negative if resources are released during an execution.
\item We develop a quantitative Hoare logic for interactively deriving
  resource bounds for Clight programs.  The logic is implemented and
  proved sound in the Coq Proof Assistant with respect to the cost
  semantics of Clight.
\item We describe an automatic static analysis that computes bounds
  together with derivations in the quantitative Hoare logic.  The
  analysis is based on type-based amortized resource analysis for
  functional programs~\cite{Jost03,HoffmannAH12} and computes
  derivations by generating simple linear constraints that can be
  solved by an off-the-shelf LP solver.
\item We show with a publically available prototype implementation and
  an experimental evaluation that our novel amortized resource
  analysis works precisely and efficiently for common loop and
  recursion patterns and that the derived constant factors in the
  bounds are close to or identical with the optimal ones.
% \item We prove the completeness of the program logic
%   with respect to the operational cost semantics.
\end{enumerate}
%
Our starting point is a recent work~\cite{veristack14} in which we
have formally verified bounds on the stack usage of C programs.  A key
part of this stack-bound verification is based on a quantitative Hoare
logic for deriving abstract stack bounds for CompCert Clight programs
that depend on the depth of (recursive) function calls.  % For
% non-recursive programs it is possible to compute derivations for such
% stack bounds in the quantitative Hoare logic automatically.  
By compiling these bounds together with the C program, we derive
verified stack bounds for x86 assembly code.

Our \emph{first contribution} is a generalization of the quantitative
Hoare logic for CompCert Clight that enables us to derive resource
bounds that are parametric in the resource of interest.  The
quantitative logic is implemented in Coq and proved sound with respect
to a small-step operational cost semantics for Clight that counts a
constant resource consumption for each atomic execution step.  The
constant resource consumption at each step is parametric in
user-definable cost metric and can be negative to express the release
of resources (as needed, e.g., to reason about memory usage).
Furthermore, a specific \emph{tick} function can be inserted in the
source code to further customize the cost model.  In the
implementation of the logic, we use a shallow embedding in Coq which
makes the logic very flexible.  There is practically no limitation on
the format of the derived resource bounds since they can be any
function that is definable in Coq.  The derived bounds can also be
parametric in a set of cost metrics or specifically apply to one fixed
cost metric.

The precision and flexibility of the quantitative Hoare logic provide
a ample foundation for reasoning about resource consumption of Clight
programs.  The logic is naturally compositional and the quantitative
Hoare triples can be used to specify resource cost of library
functions or to describe resource contracts between different
functions.  However, reasoning about quantitative properties can often
be more tedious then reasoning about intentional properties.
Consequently, automation is inevitable to derive resource bounds for
large code bases like CertiKOS.  Such an automation was easily
achievable in the case of constant stack bounds for a code base that
does not contain recursive functions.  In general however, resource
bounds have to be parametric in the arguments of a function and depend
on the number of loop iterations and (recursive) function calls that
are performed by the function.

For functional programs, there exist resource analysis systems that
are based on type-based amortized resource analysis~\cite{Jost03,
  HoffmannAH12} and can automatically derive complex polynomial bounds
with tight constant factors.  Type-based amortized resource analysis
has been our main inspiration in the design of the quantitative Hoare
logic and is therefore a natural candidate to automate the reasoning.
However, it is a well-known problem that type-based amortized resource
analysis works well for functional programs with pattern matching but
not for C-like programs with control flow that depends integer
arithmetic.  For example, it is a long-time open problem how to use
type-based amortized analysis to derive bounds for functions whose
resource consumption depends on (possibly negative) integers and
non-sequential control flow as introduced by \code{break} and
\code{return} statements.  Unfortunately, these are the functions with
non-constant resource consumption that are most common in CertiKOS and
other system software.

Our \emph{second contribution} is an automatic amortized resource
analysis for CompCert Clight that computes derivations in the
quantitative Hoare logic.  It is the first automatic amortized
resource analysis that can derive bounds that depend on negative
integers and handles programs with mutually recursive as well as
\code{break} and \code{return} statements.  It targets programs that
are common in system software, that is, programs whose resource
consumption can be described as a function of sizes $|[x,y]| =
\max(0,y-x)$ of intervals of integer variables.  Such bounds arise for
instance from standard \emph{for loops} \code{for (i = x; i+K \leq y;
  i=i+K)} where the step-size $K>0$ is a constant.

The use of interval sizes in potential functions (an therefore in
resource bounds) is the main innovation that makes the analysis
practical for C.
%
Despite the apparent simplicity of the new analysis system, it is able
to reproduce results from the literature (e.g,
SPEED~\cite{GulwaniMC09}) that have been obtained using sophisticated
abstract--interpretation base methods.  In contrast with abstract
interpretation--based methods, our technique does not require any
fixpoint computations to obtain loop invariants.  The mechanism we
designed is able to leverage local assertions such as $x < y$ that we
collect along the branching points of the program to obtain global
resource invariants.  We achieve this by generating a linear
constraint system that reflects resource cost and size changes of
integer variables in the program.  A solution of the linear program is
a resource usage bound for the C program.  % We can naturally, handle
% advanced control flow such as break, return, iteration over negative
% integer variables, and mutually-recursive functions.

@@

To the best of our knowledge, this article presents the first resource
analysis framework for C that makes it possible to combine non-trivial
automatically derived bounds with interactively derived bounds in a
prove system that produces verifiable certificates for the bounds.
Our approach complements existing work since it provides a semantic
foundation for the computation of bounds while still providing support
for automation that sometimes goes beyond the capabilities of existing
techniques.  Both the automatic amortized analysis and the
quantitative logic are naturally compositional and describe the
resource behavior of code fragments without referring to the source
code.  As a result, we naturally derive global bounds that are
functions of the input parameters of the program.  Another unique
feature of our system is, that it can naturally handle resources like
memory that may become available during the execution.

@@

In this section, we informally introduce the quantitative program
logic and the automatic amortized analysis for for Clight programs.
The purpose of this section is to build some intuition that should
make the following sections, which contain more details, easy to read.
